### 2021-02-26 written/eLifeRevision/Bittner_eLife2020.tex
abstract update

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or in terms of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choices of model parameters.
When analytic derivation of the relationship between model parameters and computational properties is intractable, approximate inference and simulation-based techniques are relied upon for scientific insight.
%Historically, the gold standard has been to analytically derive the relationship between model parameters and computational properties.  
%However, this enterprise quickly becomes infeasible as biologically realistic constraints are included into the model increasing its complexity, often resulting in \emph{ad hoc} approaches to understanding the relationship between model and computation.
We bring the use of deep generative models for probabilistic inference to bear on this problem, learning complex distributions of parameters that produce the specified properties of computation.
Our novel method solves the inverse problem by identifying the full space of parameters producing the emergent property.
%Importantly, the techniques we introduce offer a principled means to understand the implications of model parameter choices on computational properties of interest.  
We motivate this methodology with a worked example analyzing sensitivity in the stomatogastric ganglion.  
%We then use it to go beyond linear theory of neuron-type input-responsivity in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and attribute error to connectivity properties in recurrent neural networks solving a simple mathematical task.
We then use it to reveal the key factors of variability in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and scale inference of large low-rank RNN's exhibiting stable amplification.
This work illustrates how we can further leverage the power of deep learning towards solving inverse problems in theoretical neuroscience.
%More generally, this work suggests a departure from realism vs tractability considerations, towards the use of modern machine learning for sophisticated interrogation of biologically relevant models.

\section{Introduction}
The fundamental practice of theoretical neuroscience is to use a mathematical model to understand neural computation, whether that computation enables perception, action, or some intermediate processing.  
A neural computation is systematized with a set of equations -- the model -- and these equations are motivated by biophysics, neurophysiology, and other conceptual considerations \cite{kopell1988coupled,  marder1998biophysics, abbott2008theoretical, wang2010neurophysiological}.
The function of this system is governed by the choice of model \emph{parameters}, which when configured in a particular way, give rise to a measurable signature of a computation.   
The work of analyzing a model then requires solving the inverse problem: given a computation of interest, how can we reason about particular parameter configurations?  
%The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, attractor states and phase transitions, and predictions made by the model.
The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, and predictions made by the model \cite{gutenkunst2007universally, o2014cell}.

Consider the idealized practice: one carefully designs a model and analytically derives how computational properties determine model parameters.
Seminal examples of this gold standard (which often adopt approaches from statistical physics) include our field's understanding of memory capacity in associative neural networks \cite{hopfield1982neural}, chaos and autocorrelation timescales in random neural networks \cite{sompolinsky1988chaos}, the paradoxical effect \cite{tsodyks1997paradoxical}, and decision making \cite{wong2006recurrent}.
Unfortunately, as circuit models include more biological realism, theory via analytical derivation becomes intractable.
%This creates an unfavorable tradeoff.  On the one hand, one may tractably analyze systems of equations with unrealistic assumptions (for example symmetry or gaussianity), mathematically formalizing how parameters affect computation in a too-simple model.  
%On the other hand, one may choose a more biologically accurate, scientifically relevant model at the cost of \emph{ad hoc} approaches to analysis (such as simply examining simulated activity), potentially resulting in bad inference of parameters and thus erroneous scientific predictions or conclusions.
Alternatively, we can gain insight into these complex models by identifying the full distribution of parameters consistent with specified emergent phenomena.
By solving the inverse problem in this way, scientists can reason about the sensitivity and robustness of the model with respect to different parameter combinations \cite{foster1993significance, prinz2004similar, achard2006complex, o2014cell, alonso2019visualization}.

The preferred formalism for parameter identification in science is statistical inference, which has been used to great success in neuroscience through the stipulation of statistical generative models \cite{kass2001spike, brown1998statistical, paninski2004maximum, truccolo2005point, schneidman2006weak, druckmann2007novel, turner2007maximum, byron2009gaussian, macke2011empirical, park2011bayesian, granot2013stimulus, latimer2015single, lakshminarasimhan2018dynamic, duncker2019learning, ladenbauer2019inferring} (see review, \cite{paninski2018neural}). 
%Exact Bayesian inference is rarely tractable, so Markov chain Monte Carlo (MCMC) and variation inference (VI) techniques are used in practice.
However, most neural circuit models in theoretical neuroscience stipulate a noisy system of differential equations that can only be sampled or realized through forward simulation; they lack the explicit likelihood central to the probabilistic modeling toolkit.
Therefore, the most popular approaches to the inverse problem have been likelihood-free methods such as approximate Bayesian computation (ABC) \cite{sisson2007sequential, liepe2014framework}, in which reasonable parameters are obtained via simulation and rejection.

% now transition to deep ML...
%Of course, this same tradeoff has been confronted in many scientific fields characterized by the need to do inference in complex models.
%In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as a powerful inference engine: a flexible function family that can map observed phenomena (in this case the measurable signal of some computation) back to probability distributions quantifying the likely parameter configurations.  
Of course, the challenge of doing inference in complex models has arisen in many scientific fields.
In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as powerful inference engines: a flexible function family that can map observations back to probability distributions quantifying the likely parameter configurations. 
%One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data. 
One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder (VAE) \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data. 
Indeed, these tools have been used to great success in neuroscience as well, in particular for interrogating hidden states in models of both cortical population activity \cite{gao2016linear, zhao2017recursive, barello2018sparse, pandarinath2018inferring} and animal behavior \cite{wiltschko2015mapping, johnson2016composing, batty2019behavenet}. 
These works have used deep neural networks to expand the domain of neural data sets amenable to statistical modeling \cite{paninski2018neural}.

% What's wrong with existing methodology:
Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Without an efficient calculation of the probability of model properties given model parameters, neuroscientists resort to approximate Bayesian computation \cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential}, which requires a rejection heuristic, scales poorly, and only produces sets of accepted parameters lacking probabilities.
Second, there is an undesirable trade-off between the flexibility and sampling speed of approximated posterior distributions.
Sampling-based inference approaches (e.g. ABC and Markov chain Monte Carlo (MCMC) \cite{hastings1970monte, metropolis1953equation}) confer flexible approximations, yet scale poorly in number of parameters.
While variational inference (VI) \cite{saul1998mean} often results in fast posterior sampling, existing practice relies heavily on simplified classes of distributions \cite{rezende2015variational}.  
Third, such parameter inference methods are designed to operate on experimentally collected data-sets.
Ultimately, the objects of interest in theoretical neuroscience are phenomena or features of the model rather than singular data-sets.

@@

% now we've constructed the tension that theory models need this and statistical models know how to do this... now connect!
To address these three challenges, we developed an inference methodology -- `emergent property inference' -- which learns a distribution over parameter configurations in a theoretical model.  
This distribution has two critical properties: \emph{(i)} it is chosen such that draws from the distribution (parameter configurations) correspond to systems of equations that give rise to a specified emergent property (a set of constraints); and \emph{(ii)} it is chosen to have maximum entropy given those constraints, such that we identify all likely parameters and can use the distribution to reason about parametric sensitivity and degeneracies \cite{transtrum2015perspective}.  
%First, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2016density, papamakarios2017masked}.
%Second, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
%Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model, which is ultimately the object of interest in theoretical neuroscience.
%Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
%Third,  because we cannot assume the theoretical model has explicit likelihood on data or the emergent property of interest, we use stochastic gradient techniques in the spirit of likelihood free variational inference \cite{tran2017hierarchical}.    
%Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
%We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.
First, we use stochastic gradient techniques in the spirit of likelihood-free variational inference \cite{tran2017hierarchical} to enable inference in likelihood-free models of neural computation.
Second, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2017density, papamakarios2017masked}, which confers fast sampling and sensitivity measurements.
Third, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model.
Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to decision making), and network scale (from four to hundreds of neurons).
%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
First, we use EPI to understand the characteristics of noise across mulitple neuron-type populations that govern variability in a model of primary visual cortex.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
Then, we use EPI to infer multiple regimes of superior colliculus connectivity that perform rapid task switching.
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{litwin2016inhibitory, duan2018collicular}.
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.  
Finally, we emphasize the scalability of EPI by inferring high-dimensional distributions of RNNs exhibiting stable amplification.
These results point to the value of deep inference for the interrogation of biologically relevant models.

%-------------------

>>>>

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or in terms of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choices of model parameters.
%When analytic derivation of the relationship between model parameters and computational properties is intractable, approximate inference and simulation-based techniques are relied upon for scientific insight.
%We bring the use of deep generative models for probabilistic inference to bear on this problem, learning complex distributions of parameters that produce the specified properties of computation.
We present a novel technique designed to infer parameter distributions that produce mathematically defined emergent properties of computation, rather than sets of experimental data.
We find that conditioning parameter distributions on an emergent property rather than datasets is natural and often advantageous in theoretical practice.
Also, this approach allows us to greatly simplify the deep inference architecture used and to scale inference to high dimensions.
We bring this deep inference technique to bear on important questions in theoretical neuroscience, and demonstrate the broad range of insightful analyses this approach allows.
First, we show a worked example analyzing parametric sensitivity in the stomatogastric ganglion, emphasizing the general applicability of this approach to complex models with complex emergent properties.
Then, in a model of primary visual cortex where mathematical analysis becomes untenable, we identify a nonlinear manifold revealing the rule of parametric degeneracy with respect to variability.
Next, we gain a mechanistic understanding of how superior colliculus connectivity produces rapid task switching in multiple parametric regimes.
Finally, we scale inference of RNN connectivity exhibiting stable amplification to 1,000-dimensional parameter spaces.
These analyses illustrate how we can further leverage the power of deep learning towards solving inverse problems in theoretical neuroscience.

%-------- old sentences -----------

%Historically, the gold standard has been to analytically derive the relationship between model parameters and computational properties.  
%However, this enterprise quickly becomes infeasible as biologically realistic constraints are included into the model increasing its complexity, often resulting in \emph{ad hoc} approaches to understanding the relationship between model and computation.
%Importantly, the techniques we introduce offer a principled means to understand the implications of model parameter choices on computational properties of interest.  
%We then use it to go beyond linear theory of neuron-type input-responsivity in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and attribute error to connectivity properties in recurrent neural networks solving a simple mathematical task.
%More generally, this work suggests a departure from realism vs tractability considerations, towards the use of modern machine learning for sophisticated interrogation of biologically relevant models.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
The fundamental practice of theoretical neuroscience is to use a mathematical model to understand neural computation, whether that computation enables perception, action, or some intermediate processing.  
A neural computation is systematized with a set of equations -- the model -- and these equations are motivated by biophysics, neurophysiology, and other conceptual considerations \cite{kopell1988coupled,  marder1998biophysics, abbott2008theoretical, wang2010neurophysiological}.
The function of this system is governed by the choice of model \emph{parameters}, which when configured in a particular way, give rise to a measurable signature of a computation.   
The work of analyzing a model then requires solving the inverse problem: given a computation of interest, how can we reason about particular parameter configurations?  
The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, and predictions made by the model \cite{gutenkunst2007universally, o2014cell}.

Consider the idealized practice: one carefully designs a model and analytically derives how computational properties determine model parameters.
Seminal examples of this gold standard (which often adopt approaches from statistical physics) include our field's understanding of memory capacity in associative neural networks \cite{hopfield1982neural}, chaos and autocorrelation timescales in random neural networks \cite{sompolinsky1988chaos}, the paradoxical effect \cite{tsodyks1997paradoxical}, and decision making \cite{wong2006recurrent}.
Unfortunately, as circuit models include more biological realism, theory via analytical derivation becomes intractable.
Alternatively, we can gain insight into these complex models by identifying the full distribution of parameters consistent with specified emergent phenomena.
By solving the inverse problem in this way, scientists can reason about the sensitivity and robustness of the model with respect to different parameter combinations \cite{foster1993significance, prinz2004similar, achard2006complex, o2014cell, alonso2019visualization}.

The preferred formalism for parameter identification in science is statistical inference, which has been used to great success in neuroscience through the stipulation of statistical generative models \cite{kass2001spike, brown1998statistical, paninski2004maximum, truccolo2005point, schneidman2006weak, druckmann2007novel, turner2007maximum, byron2009gaussian, macke2011empirical, park2011bayesian, granot2013stimulus, latimer2015single, lakshminarasimhan2018dynamic, duncker2019learning, ladenbauer2019inferring} (see review, \cite{paninski2018neural}). 
However, most neural circuit models in theoretical neuroscience stipulate a noisy system of differential equations that can only be sampled or realized through forward simulation; they lack the explicit likelihood central to the probabilistic modeling toolkit.
Therefore, the most popular approaches to the inverse problem have been likelihood-free methods such as approximate Bayesian computation (ABC) \cite{sisson2007sequential, liepe2014framework}, in which reasonable parameters are obtained via simulation and rejection.

Of course, the challenge of doing inference in complex models has arisen in many scientific fields.
In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as powerful inference engines: a flexible function family that can map observations back to probability distributions quantifying the likely parameter configurations. 
One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder (VAE) \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data. 
Indeed, these tools have been used to great success in neuroscience as well, in particular for interrogating hidden states in models of both cortical population activity \cite{gao2016linear, zhao2017recursive, barello2018sparse, pandarinath2018inferring} and animal behavior \cite{wiltschko2015mapping, johnson2016composing, batty2019behavenet}. 
These works have used deep neural networks to expand the domain of neural data sets amenable to statistical modeling \cite{paninski2018neural}.

Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Without an efficient calculation of the probability of model properties given model parameters, neuroscientists resort to approximate Bayesian computation \cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential}, which requires a rejection heuristic, scales poorly, and only produces sets of accepted parameters lacking probabilities.
Second, there is an undesirable trade-off between the flexibility and sampling speed of approximated posterior distributions.
Sampling-based inference approaches (e.g. ABC and Markov chain Monte Carlo (MCMC) \cite{hastings1970monte, metropolis1953equation}) confer flexible approximations, yet scale poorly in number of parameters.
While variational inference (VI) \cite{saul1998mean} often results in fast posterior sampling, existing practice relies heavily on simplified classes of distributions \cite{rezende2015variational}.  
Third, such parameter inference methods are designed to operate on experimentally collected data-sets.
Ultimately, the objects of interest in theoretical neuroscience are phenomena or features of the model rather than singular data-sets.

To address these three challenges, we developed an inference methodology -- `emergent property inference' -- which learns a distribution over parameter configurations in a theoretical model.  
This distribution has two critical properties: \emph{(i)} it is chosen such that draws from the distribution (parameter configurations) correspond to systems of equations that give rise to a specified emergent property (a set of constraints); and \emph{(ii)} it is chosen to have maximum entropy given those constraints, such that we identify all likely parameters and can use the distribution to reason about parametric sensitivity and degeneracies \cite{transtrum2015perspective}.  
First, we use stochastic gradient techniques in the spirit of likelihood-free variational inference \cite{tran2017hierarchical} to enable inference in likelihood-free models of neural computation.
Second, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2017density, papamakarios2017masked}, which confers fast sampling and sensitivity measurements.
Third, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model.
Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to decision making), and network scale (from four to hundreds of neurons).
First, we use EPI to understand the characteristics of noise across mulitple neuron-type populations that govern variability in a model of primary visual cortex.
Then, we use EPI to infer multiple regimes of superior colliculus connectivity that perform rapid task switching.
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{litwin2016inhibitory, duan2018collicular}.
Finally, we emphasize the scalability of EPI by inferring high-dimensional distributions of RNNs exhibiting stable amplification.
These results point to the value of deep inference for the interrogation of biologically relevant models.

%-------- old sentences -----------

%The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, attractor states and phase transitions, and predictions made by the model.

%This creates an unfavorable tradeoff.  On the one hand, one may tractably analyze systems of equations with unrealistic assumptions (for example symmetry or gaussianity), mathematically formalizing how parameters affect computation in a too-simple model.  
%On the other hand, one may choose a more biologically accurate, scientifically relevant model at the cost of \emph{ad hoc} approaches to analysis (such as simply examining simulated activity), potentially resulting in bad inference of parameters and thus erroneous scientific predictions or conclusions.

%Exact Bayesian inference is rarely tractable, so Markov chain Monte Carlo (MCMC) and variation inference (VI) techniques are used in practice.

% now transition to deep ML...
%Of course, this same tradeoff has been confronted in many scientific fields characterized by the need to do inference in complex models.
%In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as a powerful inference engine: a flexible function family that can map observed phenomena (in this case the measurable signal of some computation) back to probability distributions quantifying the likely parameter configurations. 
%One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data.

@@

%First, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2016density, papamakarios2017masked}.
%Second, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
%Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model, which is ultimately the object of interest in theoretical neuroscience.
%Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
%Third,  because we cannot assume the theoretical model has explicit likelihood on data or the emergent property of interest, we use stochastic gradient techniques in the spirit of likelihood free variational inference \cite{tran2017hierarchical}.    
%Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
%We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.

%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.
